Penalized linear mixed models for structured genetic data

### 2021-03-04 main.tex
Adding pathwise MSE figure

To illustrate the statistical properties demonstrated by our simulations on a more realistic genome-wide scale, we present an analysis of the full Empirical genomic data ($n=588, p=526,266$) from \citet{larkin2015objectives} and a synthetic phenotype as a pseudo-real data case study. We generate the synthetic phenotype as in \eqref{eqn:sim} using 50,000 causal SNPs, a 1:1 SNR ($\eta = 0.5$), and substantial environmental effects ($\xi = 0.8$). In order to compare all methods' performance fairly, we present results for the entirety of each method's $\lambda$ path.

@@

@@

The ordinary lasso is clearly most susceptible to selecting non-causal SNPs due to their association with population structure. This corresponds to the simulation results shown in Section~\ref{sec:sel_bias}, though in Figure \ref{fig:gw_pc1_corr} we more plainly see that this bias persists across the $\lambda$ path. Likewise, PC-lasso most effectively controls this type of selection bias. This is to be expected given that we are assessing the population-informative nature of incorrectly selected SNPs via correlation with PC1, which is explicitly adjusted for in the PC-lasso method. LMM-lasso yields a substantial improvement compared to the ordinary lasso in controlling this selection bias. It should be noted that the Empirical data's RRM in Figure~\ref{fig:empirical} shows that its structure is predominantly coarse. Given our results in Section~\ref{sec:sim-fine-coarse}, we might expect a still more substantial improvement over the ordinary lasso in the analysis of more finely structured genome-wide data.

@@

Our simulations experiments reveal, unsurprisingly, that PC-lasso and LMM-lasso outperform the ordinary lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-lasso may hinder estimation accuracy. Additionally, we find that LMM-lasso consistently outperforms PC-lasso, especially when the population structure is more complex. These results indicate that LMM-lasso is a more robust method in practice when the nature of the population structure is unknown.

\anna{Add discussion of genome-wide results}

@@

>>>>

To illustrate the statistical properties demonstrated by our simulations on a more realistic genome-wide scale, we present an analysis of the full Empirical genomic data ($n=588, p=526,266$) from \citet{larkin2015objectives} and a synthetic phenotype as a pseudo-real data case study. We generate the synthetic phenotype as in \eqref{eqn:sim} using 50,000 causal SNPs, a 1:1 SNR ($\eta = 0.5$), and substantial environmental effects ($\xi = 0.8$). In order to compare all methods' performance fairly and completely, we present results for the entirety of each method's $\lambda$ path.

@@

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/genomewide_mse.png}
\caption{Pathwise MSE from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
\label{fig:gw_mse}
\end{figure}

@@

The ordinary lasso is clearly most susceptible to selecting non-causal SNPs due to their association with population structure. This corresponds to the simulation results shown in Section~\ref{sec:sel_bias}, though in Figure \ref{fig:gw_pc1_corr} we more plainly see that this bias persists across the $\lambda$ path. Likewise, PC-lasso most effectively controls this type of selection bias. This is to be expected given that we are assessing the population-informative nature of incorrectly selected SNPs via correlation with PC1, which is explicitly adjusted for in the PC-lasso method. LMM-lasso yields a substantial improvement compared to the ordinary lasso in controlling this selection bias.

@@

Our simulation experiments reveal, unsurprisingly, that PC-lasso and LMM-lasso outperform the ordinary lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-lasso may hinder estimation accuracy. Additionally, we find that LMM-lasso consistently outperforms PC-lasso, especially when the population structure is more complex. These results indicate that LMM-lasso is a more robust method in practice when the nature of the population structure is unknown.

@@

\anna{The results of a genome-wide analysis of Empirical data agree with the findings of our simulation experiments, in particular that PC-lasso and LMM-lasso are able to substantially reduce selection bias compared to the ordinary lasso. Although PC-lasso was found to control this selection bias most effectively, it should be noted that because the Empirical data consists of three distinct ethnic groups, its RRM (Figure~\ref{fig:empirical}) demonstrates predominantly coarse population structure. Given our results in Section~\ref{sec:sim-fine-coarse}, we might expect LMM-lasso to provide greater improvement over the ordinary lasso in the analysis of more finely structured genome-wide data.}
